---
title: "An Introduction to Tidyverse in Data Analysis"
description: |
  A tutorial on how to use R's tidyverse package to make data tidy. Tidying data makes for a nice, clean dataframe to use in analysis. Tidyverse contains packages such as dplyr, which will be covered in this tutorial, for handling messy data wrangling in really good ways.
author:
  - name: Ben Peterson 
    url: https://runnaboibp.github.io/fleetGeekBlog
date: "`r Sys.Date()`"
output: 
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
library(tidyverse)
library(ggplot2)
library(readxl)
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction--Why the Need for Tidyverse?

Data Analysis is a tough job, there are a lot of hurdles that can get in the way of meaningful studies of datasets. R has several packages to deal with these complications, one of them being the **tidyverse** package. 

**Tidyverse** has many useful tools to clean up data and make it much easier to work with, one of which is called the **dplyr** package. Sometimes, there is a need to change variable names or group data, or make new columns with variables that you are needing to find, even subsetting your data and selecting the pieces you want as opposed to having to navigate the entire dataframe. **Dplyr** makes these things possible, and relatively simple to use. Here we will explore a few of these packages inside of the **dplyr** package. After tidying the data, **tidyverse** allows you to use a tool called a pipe operator `%>%` which allows you to feed your dataset directly into your functions, so you don't have to worry about giving each function the data argument. With thses tools, it becomes much easier to run statistical analysis tests on it and/or graphical analysis using the **ggplot2** package. 

### Background Information

BIO314--Ecology and Evolution section A uses RStudio in the lab portion of the class for data analysis, however most of their data wrangling is done using basic R and long, complex R commands. BIO314-B uses Excel for most of their data wrangling and analysis. I am currently in section A of the class, but due to scheduling conflicts I have to take section B of the lab. Thus, I was inspired to take this project on because of the disconnect between the two classes. The goal of this project was to create a tutorial in tidyverse to open an easier pathway to data analysis in the class. Ultimately, I hope that this tutorial will help in simplifying the tidying and analysis processes and eventually unify the two sections so there is one, easy way for classes to manage their data. 

## **tidyverse**--A Package Made to Clean Data

### Variable Names-- Legality 

When working with dataframes in R, it is important to remember that there are certain ways to name variables. R does not allow spaces or characters such as parentheses or forward slash.

In BIO314, one of the datasets used is from the lab on gall fly larvae. The lab involves going out into a field and collecting galls (woody growths on goldenrod plants that larvae grow in) and taking size measurements to determine what the fate of the larva was. The class data was collected in a document on Microsoft Excel.

The following code chunk will import the dataset into RStudio and set the data types for each variable so we can tidy it up. I will be importing the data into two different datasets so we can show the difference between other methods versus **tidyverse** methods.

#### Gall Data A

```{r}
A <- read_excel("studentGallData.xlsx", 
    col_types = c("text", "numeric", "numeric", 
        "numeric", "text", "text", "text"))
```

#### Gall Data B

```{r}
B <- read_excel("studentGallData.xlsx", 
    col_types = c("text", "numeric", "numeric", 
        "numeric", "text", "text", "text"))
```

First, we'll look into how to change the illegal column names. Dataframe A will be done using the `colnames()` function in the **base** R package. Dataframe B will be done using the `rename()` function in the **dplyr** package from **tidyverse**.

#### Dataframe A

First, using `colnames()` from the **base** package, we can change each illegal variable name individually in separate lines. The function takes the illegal column name in quotes inside the brackets and replaces it with the desired column name as an assignment. 

The command would look like:

```{r eval=FALSE}
colnames(data)[colnames(data)=="Illegal Name"] <- "newVarName"
```

Now, an example using our newly created A dataset of gall data. Note that this must be done for each individual variable name you want to change. 

```{r}
colnames(A)[colnames(A)=="Gall Ht (cm)"] <- "gallHeightA"
colnames(A)[colnames(A)=="Gall Dia (mm)"] <- "gallDiameterA"
colnames(A)[colnames(A)=="Hole Dia (mm)"] <- "holeDiameterA"
colnames(A)[colnames(A)=="Expected Cause (EG, WP, C)"] <- "expectedCauseA"
colnames(A)[colnames(A)=="A / N"] <- "awayOrNearA"
```

```{r}
DT::datatable(A, options = list(scrollX = TRUE))
```

As you can see, this gets the job done okay, but it is messy to look at and pretty complex. Also, this takes five separate functions to do, so it is rather tedious. Surely, there has got to be a better way. 

#### Dataframe B

Let's take a look at how the `rename()` function in the **dplyr** package can make this a lot more simple, and better yet, involve only one, easy to understand function.

**dplyr**'s `rename()` function is easy! It takes the dataset you're using as the first argument and then you simply give your name change in the format shown:

```{r eval=FALSE}
df <-
rename(df, newVar1 = `Old Illegal Var`,
           newVar2 = `Second Illegal`)
```  
           
The backticks show R that you might be feeding it illegal variable names and so it looks for those too. To save the dataframe with the new variables, simply use the assignment operator to save it. In this example, I wil take our B dataset and name it tidyB since it has the nicer, legal names.  

```{r}
tidyB <-
rename(B, gallHeightB = `Gall Ht (cm)`,
          gallDiameterB = `Gall Dia (mm)`,
          holeDiameterB = `Hole Dia (mm)`,
          expectedCauseB = `Expected Cause (EG, WP, C)`,
          awayOrNearB = `A / N`)
```

```{r}
DT::datatable(tidyB, options = list(scrollX = TRUE))
```

Clearly, this also gets the job done, but in a much cleaner, easier to understand way.

### Selecting varaibles to work with using **dplyr**

When working with data in R, very rarely will you want to work with the entire dataset at once. Often you will be looking at specific varaibles to search for patterns and statistical significance. This means that you will want to subset your data.

#### Base R Method

This can be done using the **base** R package, but it is tedious and requires you to remember which column numbers your desired variables are in. You'd need to call your dataset and then use brackets--you'd structure the command with your row numbers,column numbers. So you'd have data[rows,columns]. If you only want to use columns, then you'd simply leave the rows argument blank and just have column numbers after the comma inside a c() command. 

This command would look like:

```{r eval=FALSE}
data <-
data[,c(col1, col2, col3, ... )]
DT::datatable(data)
```


```{r}
A<-
A[,c(2,3,4,5,6,7)]
DT::datatable(A, options = list(scrollX = TRUE))
```

Like before, the **base** R method gets the job done, but can be quite tedious depending on what your dataset looks like. Also, this requires you to remember column numbers and you might accidentally select wrong variables.

#### Tidyverse -- **dplyr**'s `select()` function

Like before, **tidyverse** includes a much simpler method to subsetting data, especially after you've given your data the legal names with the `rename()` function. This can be achieved using the `select()` function. Here, you will pipe your dataset in and just name the variables desired separated by commas. This looks relatively simple, and since you gave the dataset the nice names, it should be easy to remember. I'll pipe this into a DT::datatable so the output is nicer. 

The command would have the form:

```{r eval=FALSE}
data %>%
  select(var1, var2, var3, ... ) %>%
  DT::datatable() #If you wanted to make a prettier table
``` 

Now, an example of this using our tidyB gall dataset:

```{r}
tidyB <-
tidyB %>%
  select(gallHeightB, gallDiameterB, holeDiameterB, expectedCauseB, awayOrNearB)
```

```{r}
tidyB %>%
DT::datatable(options = list(scrollX = TRUE))
```


This looks much nicer, and is a more user-friendly way of subsetting data.

### Grouping Variables using **dplyr**'s `group_by()` function

Sometimes, it is necessary to group data by a certain variable to compare and contrast. This can be done easily using **dplyr**'s `group_by()` function! Simply pipe your data into the `group_by()` function and give the variable name you want to group by as your argument. Then, you can run analysis on it to compare means, values, etc.

The command would look like: 

```{r eval=FALSE}
data %>%
  group_by(varName)
```

I'll show an example of this with the tidyB gall dataset plotting the number of cases for each group of expected gall hole causes in a bar graph.  

```{r preview=TRUE}
tidyB %>%
  group_by(expectedCauseB) %>%
  ggplot(aes(x=expectedCauseB)) +
    geom_bar(aes(fill=expectedCauseB)) + 
    labs(x="Expected Cause of Hole found in Gall",
         y="Count of Occurances",
         title="Count of Expected Causes of Holes in Galls created by the 
                Goldenrod Gall Fly")
```


### Summarizing data using **dplyr**

Often, you will find that you wish to find a summary of values in a dataset to use in analysis, such as a mean value. **Dplyr** has a handy tool for this in the `summarize()` function, which creates a new summary variable in the dataset to return the desired information. The `summarize()` function takes an argument in the form of the new variable name being equal to the value that you wish to assign this new variable.

Now, I'll use the `summarize()` function to find the mean gall height for each of the groups of expected causes. Note that you can do more than one summary variable at a time, simply separate by commas as shown in this example. Also, the 'na.rm=TRUE' argument is a way to let R know to ignore NA values in the counts where an expected cause was identified.

```{r}
tidyBSummary <-
tidyB %>%
  group_by(expectedCauseB) %>%
  summarize(meanDia = mean(gallDiameterB, na.rm=TRUE),
            meanHt = mean(gallHeightB, na.rm=TRUE),
            meanHoleDia = mean(holeDiameterB, na.rm=TRUE))
```

```{r}
tidyBSummary %>%
  knitr::kable(caption="Mean Gall Diameter, Gall Height, and Gall Hole diameter for the
               five possible causes. Note `NA` represents a Gall where the gall fly 
               larva was still alive inside the gall.")
```




